 

### 2020-08-19 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.** 
Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--typically T, PPT, and PDSI--over the current year and, typically, previous year. 
While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS), this approach does not systematically examine the numerous potential multiple-month time windows to determine the single most influential window over which a given variable influences annual growth. 
Needed is an objective approach to systematically evaluate the numerous potential climate variables and time windows. 
Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; *but see VS-Lite* @foster_predicting_2016]. 
Such effects are expected based on observations that temperature and moisture interactively shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @taylor_temperature_2017; EXTRATROPICAL_AND_TREE_LEVEL_REFS; @banbury_morgan_global_nodate].

@@

We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015], plus one in New Mexico, representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
Trees were cored within or close to (max distance: # km) the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies [@vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; MORE]. 
In using this variety of data sources, we encountered and solved a variety of challenges for analyzing existing tree-ring records with this approach.

@@

All tree cores were measured and cross-dated using standard dendrochronological practices. 
The full record for all cores was retained for analysis, with the following exceptions.
First, we excluded species with <7 cores.
Second, we excluded cores with <30 years of record.
Third, for any sites with a record of **$\Delta r$ > 10 mm **within the first 15 years of the record, we excluded those 15 years.
Fourth, for trees cored dead, we excluded the final 20 years prior to death to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
Finally, for each species, we excluded any reconstructed $DBH$ range represented by <3 cores.

@@

Monthly climate data for 1901-20XX were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
Variables considered here included mean, minimum, and maximum temperatures; potential evapotranspiration (PET); precipitation; and wet day frequency. 
While the CRU database has records for every year, some have been gap-filled based on long-term monthly means [@harris_updated_2014; @harris_version_2020]. 
To ... (**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
(*Describe criteria for excluding variables/ time frames.*)
(*Describe BCI data.* )

@@

First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation). 
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a generalized least squares (GLS) model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))

We then used *climwin* to identify the best climate variables across the full set of cores from each site, noting that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (Table #). (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51))
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models.

A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek (*and NM*?). 
There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most significant climate drivers.
(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?**)

@@

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers. Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each. The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-Aug; black boxes), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where PET is one of several driver variables (*specify model*). Plotted are responses of species for which PET was identified as a signficant driver in the top model.](tables_figures/quilt_comparison.png)

**Across the three metrics of growth, the "landscape" of climate effects over various time windows was generally similar, but the optimal time window or even the top climate variable sometimes differed (Figs. S2-S3).**
Specifically, $\Delta r$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups. 
In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. S2). 
In #  cases, *climwin* identified the same climate variable but different time windows. 
In # cases, *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. S3).
Finally, in # cases, *climwin* identified different climate variables and different time windows. 
Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric.

@@

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in the precipitation and temperature variable groups. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

Precipitation responses were observed at all sites and *for all but # species* (Fig. 3).
Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
Responses were most commonly positive and were most pronounced at the driest site (Little Tesque). 
...
Responses commonly showed some non-linearity, and in some cases it was quite pronounced (e.g., Little Tesque), with the most common pattern being a decelerating increase *([ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.

Temperature responses were observed at all sites *(but BCI? [ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig. 3).
Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
*(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
*(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
Responses commonly showed some non-linearity, but there were *few, if any* cases of pronounced nonlinearity *(but how do we define this? [ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.

@@

On climate sensitivity:
- Climwin step is problematic when climate is rapidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).

@@

Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014], we found evidence of saturation or decline in the majority **(`r round(20/26*100, digits=0)`%)** of species-site combinations analyzed.
 *(check Foster et al. 2006)*
$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

>>>>

**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.** 
Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--typically T, PPT, and PDSI--over the current year and, typically, previous year. 
While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS), this approach does not systematically examine the numerous potential multiple-month time windows to determine the single most influential window over which a given variable influences annual growth. 
Needed is an objective approach to systematically evaluate the numerous potential climate variables and time windows. 
Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; *but see VS-Lite* @foster_predicting_2016]. 
Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @taylor_temperature_2017; EXTRATROPICAL_AND_TREE_LEVEL_REFS; @banbury_morgan_global_nodate].

@@

We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables 1, S1-S2).
Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015].
Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Table 1). 
In using this variety of data sources, we ensured that our approach was able to handle a variety of challenges presented by varying methodologies and forest types.

@@

All tree cores were measured and cross-dated by the original researchers using standard dendrochronological practices. 
The full record for all cores was retained for analysis, with the exception of records that had to be excluded due to insufficient sample size or anomalous growth patterns.
Specifically, we excluded
(1) species with <7 cores,
(2) cores with <30 years of record,
(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction, 
and (4) the final 20 years prior to death of trees cored dead. 
The final criteria was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
From analyses including $DBH$ (see below), we further excluded
(1) trees for which we lacked data required to reconstruct $DBH$,
(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $\Delta r$'s across the core (*Appendix S1*),
(3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.

@@

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); potential evapotranspiration ($PET$); precipitation ($PPT$); and precipitation day frequency ($PDF$). 
For BCI, we calculated $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0-alpha [@anderson-teixeira_forestgeoclimate_2020].

@@

First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation). 
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a generalized additive model (GAM) to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))

We then used *climwin* to identify the best climate variables across the full set of cores from each site, noting that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*). (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51))
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models.

A challenge to this system arose for the sites that have undergone the most rapid changes in climate and tree growth: SC and LT, where trees exhibit significant growth declines attributed to rising temperatures [@sniderhan_growth_2016] and increasing drought (REF), respectively. 
*([ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis (Appendix S3).
After determining that... , here we present results *(method)*.

@@

*Testing for climate-DBH interactions*

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers. Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each. The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model). Plotted are responses of species for which $PET$ was included in the top model.](tables_figures/quilt_comparison.png)

**Across the three metrics of growth, the "landscape" of climate effects over various time windows was generally similar, but the optimal time window or even the top climate variable sometimes differed (Figs. S2-S4).**
Specifically, $\Delta r$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups. 
In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. S2). 
In #  cases, *climwin* identified the same climate variable but different time windows. 
In # cases, *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. S3).
Finally, in # cases, *climwin* identified different climate variables and different time windows (e.g., Fig. S4).
Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)

@@

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in the precipitation and temperature variable groups. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. THIS FIGURE STILL NEEDS 2 MORE SITES... WONDER IF THERE'S A BETTER LAYOUT...](tables_figures/pre_temp_groups.png)

@@

Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. 3).
Responses were most commonly positive, and were most pronounced at the driest site (Little Tesque). 
Responses commonly showed some non-linearity, and in some cases it was quite pronounced (e.g., Little Tesque), with the most common pattern being a decelerating increase *([ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.

Temperature responses were included in the best model at all sites *(but BCI? [ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig. 3).
Specifically, a temperature term was included in the best model for # of # site-species combinations.
Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
*(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
*(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
Responses commonly showed some non-linearity, but there were *few, if any* cases of pronounced nonlinearity *(but how do we define this? [ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.

@@

Ideas to discuss:

- temperature sensitivity shifts from neg in warm climates to positive in cold climates, although @sniderhan_growth_2016 shows that the effect shifted to negative as warming progressed

- additive effects are common and should not be overlooked

- nonlinear effects occur sometimes

@@

Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014] and has also been observed in tree-rings [@foster_predicting_2016], we found evidence of saturation or decline in the majority **(`r round(20/26*100, digits=0)`%)** of species-site combinations analyzed.
$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

### 2020-08-19 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript

## Appendix S2. Methods for comparing 
(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE, caption = "Table S1. List of species analyzed", escape = F)%>%
  kable_styling(latex_options = c("scale_down", "hold_position")) 
```

@@

** Table S2- allometric equations for bark thickness **

@@

>>>>

## Appendix S2. Methods for comparing climwin results with traditional methods
(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)

@@

## Appendix S3. Dealing with rapidly changing climate and tree growth

[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25)

Our analysis included two sites where climate change has had pronounced effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and Little Tesque, New Mexico, USA (LT).
At SC, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("tables_figures/site_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE,  caption = "Table S1. Site details", escape = F)%>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

\newpage
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S2 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S2, format = "latex", booktabs = TRUE, caption = "Table S2. List of species analyzed", escape = F)%>%
  kable_styling(latex_options = c("scale_down", "hold_position")) 
```

** Table S3- allometric equations for bark thickness **

@@

![**Figure S2 | Example comparison of climwin output when the optimal climate variable and time window were identical across growth metrics.** Example is temperature variable group at Little Tesque, NM. Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($\Delta r$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($\Delta r$-trees with $DBH$, BAI, $\Delta AGB$).)](tables_figures/NewMexico_pet.png)

![**Figure S3 | Example comparison of climwin output when the optimal climate variable differed across growth metrics, but the selected time window was similar.** Example is precipitation variable group at Little Tesque, NM. Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET) as the strongest climate variable for $\Delta AGB$.](tables_figures/NewMexico_wet_pre.png)

![**Figure S4 | Example comparison of climwin output when both the optimal climate variable and selected time window differed across growth metrics.** Example is precipitation variable group at SCBI, VA, USA. Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET)  as the strongest climate variable for $\Delta AGB$. The optimal time window (circled) also differed across growth metrics and depending on whether or not trees without DBH were included in the analysis ($\Delta r$-all trees and $\Delta r$-trees with $DBH$, respectively). ](tables_figures/SCBI_pre_wet.png)

### 2020-08-19 doc/manuscript/tables_figures/sites_table.Rmd
worked on manuscript

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
  column_spec(1, width = "2cm") %>%
  column_spec(2, width = "4cm") %>%
  column_spec(3, width = "3cm") %>%
  column_spec(6, width = "2cm") %>%
  column_spec(9, width = "2cm") %>%
  column_spec(10, width = "2cm") %>%
  column_spec(11, width = "3cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) 
```

>>>>

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
  column_spec(1, width = "1cm") %>%
  column_spec(2, width = "3cm") %>%
  column_spec(3, width = "3cm") %>%
  #column_spec(6, width = ".5cm") %>%
  column_spec(7, width = "1.4cm") %>%
  column_spec(8, width = "1cm") %>%
  #column_spec(9, width = "2cm") %>%
  column_spec(10, width = "1.4cm") %>%
  column_spec(11, width = "1.4cm") %>%
  column_spec(12, width = "3cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```
